---
title: "02 — Binary logit on sessions (+ power analysis)"
subtitle: "demand-sim · docs/03 §2, §9 · the simulator's native estimand"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
```

With session-level offers (price shown + buy/no-buy, including non-buyers),
demand is a purchase probability: `P(buy) = sigma(a + b*price)`. Identification
is clean by construction — prices are randomized across cells, so `Cov(P, u) = 0`.
The remaining lie is aggregation: the population is a discrete mixture of
segments, so a single pooled `b` is a misspecified compromise. The oracle
quantifies exactly how much.

# 1. Data

```{r data}
library(ggplot2)
library(pwr)

sessions <- read.csv("../data/fact_session.csv.gz")   # read.csv handles .gz
truth_curve <- read.csv("../data/ground_truth_curves.csv")  # ORACLE
truth_params <- read.csv("../data/truth_params.csv")        # ORACLE

# pandas writes booleans as "True"/"False" — convert to logical explicitly
sessions$purchased <- sessions$purchased == "True"
sessions$promo_flag <- sessions$promo_flag == "True"

nrow(sessions); table(sessions$cell); mean(sessions$purchased)
```

# 2. Train

```{r train}
fit <- glm(purchased ~ quoted_price + promo_flag,
           data = sessions, family = binomial())
summary(fit)$coefficients
b_hat <- coef(fit)["quoted_price"]
```

# 3. Validate against the oracle

Two checks. First, the raw cell means — the model-free demand curve the RCT
buys us — against true `P_aggregate` with 95% CIs (coverage should be ~95%).
Second, the pooled `b` against the share-weighted true `price_coef` — the gap
is the documented aggregation bias, not an error to fix.

```{r validate}
cells <- aggregate(purchased ~ quoted_price, sessions,
                   function(x) c(n = length(x), conv = mean(x)))
cells <- do.call(data.frame, cells)
names(cells) <- c("price", "n", "conv")
cells$se <- sqrt(cells$conv * (1 - cells$conv) / cells$n)
cells <- merge(cells,
               truth_curve[, c("price", "P_aggregate")], by = "price")
cells$covered <- abs(cells$conv - cells$P_aggregate) <= 1.96 * cells$se
cells

b_true <- truth_params$value[truth_params$param == "price_coef_share_weighted"]
c(b_hat = unname(b_hat), b_true_share_weighted = b_true,
  pct_gap = 100 * (b_hat - b_true) / abs(b_true))
```

Cluster note (CLAUDE.md): with ~3 sessions per consumer, session rows are not
independent. Randomization here is per-session so the point estimate is fine,
but honest SEs cluster on `consumer_id`:

```{r cluster-se}
library(sandwich); library(lmtest)
coeftest(fit, vcov = vcovCL(fit, cluster = sessions$consumer_id))[2, ]
```

# 4. The demand curve it implies

```{r curve}
grid <- data.frame(quoted_price = seq(5, 120, 1), promo_flag = FALSE)
grid$fitted <- predict(fit, newdata = grid, type = "response")

ggplot() +
  geom_line(data = truth_curve, aes(price, P_aggregate, color = "true curve (oracle)"),
            linewidth = 1) +
  geom_line(data = grid, aes(quoted_price, fitted, color = "pooled logit"),
            linetype = "dashed", linewidth = 1) +
  geom_pointrange(data = cells, aes(price, conv,
                                    ymin = conv - 1.96 * se, ymax = conv + 1.96 * se,
                                    color = "cell means (RCT)")) +
  labs(x = "price", y = "P(buy)", color = NULL,
       title = "Cell means trace the true curve; the pooled logit smooths through it")
```

The mixture makes the true curve non-logistic: pooled logit is too steep in the
middle, too flat in the tails. The cell means (model-free) hug the truth — the
argument for experiments over functional form.

# 5. Power analysis (docs/04 §2)

How many sessions per cell to detect a realistic price effect?

```{r power}
p1 <- 0.10; p2 <- 0.085     # +20% price at elasticity ~ -0.75
pwr.2p.test(h = ES.h(p1, p2), sig.level = 0.05, power = 0.80)
```

~6k per arm for one contrast; a 10-cell curve needs 60k+ before segment-level
curves multiply it again. This is why the session file has 250k rows.

# 6. Forecast (Phase 5)

Experiment-estimated price response, carried into a panel forecast: seasonal
baseline from the daily panel, scaled by the logit's relative conversion at the
posted future price. Grade against oracle demand.

```{r forecast}
panel <- read.csv("../data/fact_sales_daily.csv")
panel$date <- as.Date(panel$date)
panel$series <- paste(panel$product_id, panel$store_id, sep = "_")
cutoff <- max(panel$date) - 27
train <- subset(panel, date < cutoff)
test  <- subset(panel, date >= cutoff)

# baseline: mean sales by series x weekday (raw seasonal profile)
train$dow <- format(train$date, "%u"); test$dow <- format(test$date, "%u")
base <- aggregate(units_sold ~ series + dow, train, mean)
names(base)[3] <- "base_units"
test <- merge(test, base, by = c("series", "dow"), all.x = TRUE)

# price scaling from the session logit (price response only, rescaled to
# each series' typical price so the experiment elasticity transfers)
ref_price <- aggregate(avg_price ~ series, train, median)
names(ref_price)[2] <- "ref_price"
test <- merge(test, ref_price, by = "series")
conv <- function(p) plogis(coef(fit)[1] + coef(fit)["quoted_price"] * p)
test$pred <- test$base_units * conv(test$avg_price) / conv(test$ref_price)

err_d <- test$pred - test$units_demanded   # ORACLE target
err_s <- test$pred - test$units_sold
data.frame(target = c("units_demanded (oracle)", "units_sold"),
           mae  = c(mean(abs(err_d)), mean(abs(err_s))),
           rmse = c(sqrt(mean(err_d^2)), sqrt(mean(err_s^2))),
           bias = c(mean(err_d), mean(err_s)))
```

A crude forecaster — but the price adjustment comes from *randomized* variation,
which is the one ingredient the panel-only models in 01/06 can never certify.
