---
title: "03 — Mixed logit (random coefficients)"
subtitle: "demand-sim · docs/03 §3 · a continuous mixture chasing discrete segments"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
```

The ground truth is a finite mixture: three segments with their own price
coefficients. Mixed logit assumes the price coefficient is *continuously*
distributed (here Normal) across consumers — so this is a controlled
misspecification test: how well does a normal mixing distribution approximate
three point masses? The panel structure (repeat sessions per consumer) is what
identifies the mixing distribution at all.

Runtime note: simulated ML over 250k sessions is slow; we subsample consumers.
Estimates are noisier but the logic is identical.

# 1. Data

```{r data}
library(mlogit)
library(ggplot2)

sessions <- read.csv("../data/fact_session.csv.gz")
seg_truth <- read.csv("../data/dim_segment_truth.csv")     # ORACLE
truth_curve <- read.csv("../data/ground_truth_curves.csv") # ORACLE

# pandas writes booleans as "True"/"False" — convert to logical explicitly
sessions$purchased <- sessions$purchased == "True"
sessions$promo_flag <- sessions$promo_flag == "True"

set.seed(42)
keep <- sample(unique(sessions$consumer_id), 8000)
sub <- subset(sessions, consumer_id %in% keep)
nrow(sub)
```

# 2. Reshape to choice format

Binary choice as two alternatives: `buy` (with the quoted price) vs `nobuy`
(price 0). `mlogit` wants long format, indexed by chooser and alternative.

```{r reshape}
long <- data.frame(
  chid = rep(seq_len(nrow(sub)), each = 2),
  id   = rep(sub$consumer_id, each = 2),
  alt  = rep(c("buy", "nobuy"), nrow(sub)),
  choice = as.vector(rbind(sub$purchased, !sub$purchased)),
  price  = as.vector(rbind(sub$quoted_price, 0)),
  promo  = as.vector(rbind(as.numeric(sub$promo_flag), 0))
)
dml <- mlogit.data(long, choice = "choice", shape = "long",
                   alt.var = "alt", chid.var = "chid", id.var = "id")
```

# 3. Train

Price coefficient random Normal across consumers; Halton draws; panel = TRUE so
a consumer's repeated sessions share one draw.

```{r train, cache=TRUE}
m <- mlogit(choice ~ price + promo | 1, data = dml,
            rpar = c(price = "n"), panel = TRUE, R = 300, halton = NA)
summary(m)
```

# 4. Validate against the oracle mixture

The true mixing distribution is three point masses; compare its mean/sd to the
estimated Normal.

```{r validate}
b_mean_true <- with(seg_truth, sum(share * price_coef))
b_sd_true   <- sqrt(with(seg_truth, sum(share * (price_coef - b_mean_true)^2)))

est <- coef(m)
data.frame(
  quantity = c("mean(price coef)", "sd(price coef)"),
  mixed_logit = c(est["price"], abs(est["sd.price"])),
  truth_discrete_mixture = c(b_mean_true, b_sd_true)
)
seg_truth[, c("segment", "share", "price_coef")]
```

Read this as a misspecification result, not a failure: a Normal cannot place
45% of its mass at −0.11 and 15% at −0.012, so it stretches the sd to cover the
spread. The *mean* should be close; the shape cannot be.

# 5. The demand curve it implies

Average the logistic over the estimated mixing distribution (simulate draws of
the price coefficient), against the true mixture curve.

```{r curve}
set.seed(7)
draws <- rnorm(2000, mean = est["price"], sd = abs(est["sd.price"]))
grid <- seq(5, 120, 1)
P_hat <- sapply(grid, function(p) mean(plogis(est["(Intercept):buy"] + draws * p)))

ggplot() +
  geom_line(data = truth_curve, aes(price, P_aggregate, color = "true mixture (oracle)"),
            linewidth = 1) +
  geom_line(data = data.frame(price = grid, P = P_hat),
            aes(price, P, color = "mixed logit (normal mixing)"),
            linetype = "dashed", linewidth = 1) +
  labs(y = "P(buy)", color = NULL,
       title = "Continuous mixing approximates the discrete-segment curve")
```

The normal mixture tracks the aggregate curve far better than the pooled logit
in 02 — heterogeneity is most of what the pooled model was missing, even with
the wrong mixing family.

# 6. Forecast (Phase 5)

Same two-part pattern as 02 — seasonal baseline × relative conversion — but
with the mixed-logit curve supplying the price response.

```{r forecast}
panel <- read.csv("../data/fact_sales_daily.csv")
panel$date <- as.Date(panel$date)
panel$series <- paste(panel$product_id, panel$store_id, sep = "_")
cutoff <- max(panel$date) - 27
train <- subset(panel, date < cutoff)
test  <- subset(panel, date >= cutoff)

train$dow <- format(train$date, "%u"); test$dow <- format(test$date, "%u")
base <- aggregate(units_sold ~ series + dow, train, mean)
names(base)[3] <- "base_units"
test <- merge(test, base, by = c("series", "dow"), all.x = TRUE)
ref <- aggregate(avg_price ~ series, train, median); names(ref)[2] <- "ref_price"
test <- merge(test, ref, by = "series")

conv_mix <- function(p) mean(plogis(est["(Intercept):buy"] + draws * p))
test$pred <- test$base_units *
  sapply(test$avg_price, conv_mix) / sapply(test$ref_price, conv_mix)

err_d <- test$pred - test$units_demanded   # ORACLE target
data.frame(mae = mean(abs(err_d)), rmse = sqrt(mean(err_d^2)), bias = mean(err_d))
```

Compare with 02's table: the gain over the pooled logit comes entirely from the
heterogeneous price response bending the curve the right way at promo prices.
