---
title: "04 — Double Machine Learning (PLR vs PLIV)"
subtitle: "demand-sim · docs/03 §4, docs/05 §5 · flexible ML is not an identification strategy"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
```

The endogenous market (docs/05 §2): true elasticity θ = −1.5; the pricer sees
part of the demand shock and prices into it (γ = 0.5), so `Cov(log P, u) > 0`;
a cost shifter `z` moves price but not demand. The headline: **DML-PLR
orthogonalizes observed confounders (x) but cannot fix simultaneity — it should
land near the biased OLS. DML-PLIV, which uses the instrument, recovers θ.**

# 1. Data

```{r data}
library(DoubleML)
library(mlr3)
library(mlr3learners)   # provides regr.ranger
library(ggplot2)
lgr::get_logger("mlr3")$set_threshold("warn")

market <- read.csv("../data/endogenous_market.csv")   # observable: log_q, log_p, x, z
truth  <- read.csv("../data/truth_params.csv")
theta_true <- truth$value[truth$param == "theta_true"]
head(market); theta_true
```

# 2. Baseline: naive OLS

```{r ols}
ols <- lm(log_q ~ log_p + x, data = market)
coef(ols)["log_p"]    # expect ~ -1.1: attenuated toward zero by gamma
```

# 3. DML-PLR (no instrument) — the honest negative result

```{r plr, cache=TRUE}
set.seed(42)
dml_plr_data <- DoubleMLData$new(market, y_col = "log_q",
                                 d_cols = "log_p", x_cols = "x")
plr <- DoubleMLPLR$new(dml_plr_data,
                       ml_l = lrn("regr.ranger", num.trees = 200),
                       ml_m = lrn("regr.ranger", num.trees = 200),
                       n_folds = 5)
plr$fit()
plr$summary()
```

Random forests plus cross-fitting, and θ̂ is still ~−1.1: the ML nuisance
models absorb `x` perfectly, but the unobserved shock `u` sits in both price
and quantity, and no amount of flexible regression on observables removes it.

# 4. DML-PLIV (with the cost-shifter instrument)

```{r pliv, cache=TRUE}
set.seed(42)
dml_pliv_data <- DoubleMLData$new(market, y_col = "log_q", d_cols = "log_p",
                                  x_cols = "x", z_cols = "z")
pliv <- DoubleMLPLIV$new(dml_pliv_data,
                         ml_l = lrn("regr.ranger", num.trees = 200),
                         ml_m = lrn("regr.ranger", num.trees = 200),
                         ml_r = lrn("regr.ranger", num.trees = 200),
                         n_folds = 5)
pliv$fit()
pliv$summary()
```

# 5. Validate against the oracle

```{r validate}
res <- data.frame(
  estimator = c("OLS", "DML-PLR (no IV)", "DML-PLIV (IV)"),
  theta_hat = c(coef(ols)["log_p"], plr$coef, pliv$coef),
  se        = c(summary(ols)$coefficients["log_p", 2], plr$se, pliv$se)
)
res$bias <- res$theta_hat - theta_true
res$ci_covers_truth <- abs(res$bias) <= 1.96 * res$se
res
```

Identification strategy, named: PLIV's assumption is instrument validity —
`z` (a cost shock) shifts price and enters demand only through price. That is
true here **by construction**; in the field it is the argument you must win.

# 6. The demand curve it implies

Constant-elasticity curves through the sample mean, under each θ̂.

```{r curve}
p_bar <- mean(exp(market$log_p)); q_bar <- mean(exp(market$log_q))
grid <- seq(0.6 * p_bar, 1.6 * p_bar, length.out = 80)
curve <- rbind(
  data.frame(price = grid, q = q_bar * (grid / p_bar)^theta_true,        which = "truth (theta = -1.5)"),
  data.frame(price = grid, q = q_bar * (grid / p_bar)^res$theta_hat[1],  which = "OLS / PLR (~ -1.1)"),
  data.frame(price = grid, q = q_bar * (grid / p_bar)^res$theta_hat[3],  which = "PLIV")
)
ggplot(curve, aes(price, q, color = which)) + geom_line(linewidth = 1) +
  labs(y = "quantity", color = NULL,
       title = "The attenuated curve understates every price response")
```

# 7. Forecast (Phase 5)

Forecast the last 20% of periods with each θ̂ (nuisance g(x) refit on the
training window). The PLR/OLS curve misses hardest exactly when price moves —
which is when a pricing team consults the forecast.

```{r forecast}
n <- nrow(market); tr <- 1:floor(0.8 * n); te <- (floor(0.8 * n) + 1):n
gx <- lm(log_q ~ x, data = market[tr, ])   # crude g(x); theta applied structurally

for_theta <- function(theta) {
  pred <- predict(gx, market[te, ]) +
    theta * (market$log_p[te] - mean(market$log_p[tr]))
  err <- pred - market$log_q[te]
  c(mae = mean(abs(err)), rmse = sqrt(mean(err^2)), bias = mean(err))
}
rbind("OLS/PLR theta" = for_theta(res$theta_hat[1]),
      "PLIV theta"    = for_theta(res$theta_hat[3]),
      "true theta"    = for_theta(theta_true))
```

In-sample fit barely separates them — the endogenous DGP generates prices that
co-move with shocks, flattering the wrong θ. The separation shows up under
*intervention* (a deliberate price change), which is the forecast that matters
and the one only the causal parameter gets right. See 07 for the 2SLS/control-
function versions of the same rescue.
