---
title: "05 — Regression discontinuity at surge-rounding cutpoints"
subtitle: "demand-sim · docs/03 §5 · the Uber trick, with a checkable answer key"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
```

The pricing algorithm computes a continuous latent surge multiplier and rounds
it to 0.1 steps. Sessions just left and right of a rounding cutpoint are
statistically identical but face different prices — a discontinuity that
identifies demand *locally* with no experiment and no instrument. The firm
knows its own algorithm, so the running variable is observable. Identification
assumption: nothing else jumps at the cutpoint (no manipulation of the latent).

# 1. Data

```{r data}
library(rdrobust)
library(rddensity)
library(ggplot2)

surge <- read.csv("../data/surge_sessions.csv.gz")
truth <- read.csv("../data/truth_params.csv")
cut       <- truth$value[truth$param == "rd_cut"]         # 1.25
step      <- truth$value[truth$param == "rd_step"]        # 0.1
base_p    <- truth$value[truth$param == "base_price"]     # 25
true_jump <- truth$value[truth$param == "true_jump"]      # ORACLE

# pandas writes booleans as "True"/"False" — convert to logical explicitly
surge$purchased <- surge$purchased == "True"
surge$running <- surge$surge_latent - cut
nrow(surge); c(cut = cut, base_price = base_p, true_jump = true_jump)
```

# 2. Look first: conversion by surge level, and the rdplot

```{r look}
lev <- aggregate(purchased ~ surge_multiplier, surge, mean)
lev$price <- base_p * lev$surge_multiplier
lev

rdplot(y = surge$purchased, x = surge$running, c = 0,
       title = "Purchase probability around the 1.25 rounding cutpoint",
       x.label = "latent multiplier − cutpoint", y.label = "P(buy)")
```

# 3. Train: local-linear RD with robust bias-corrected inference

```{r train}
res <- rdrobust(y = surge$purchased, x = surge$running, c = 0)
summary(res)
```

# 4. Validate against the oracle

`rdrobust`'s conventional estimate vs the true jump P(25·1.3) − P(25·1.2) from
the demand oracle.

```{r validate}
jump_hat <- res$coef["Conventional", ]
se_hat   <- res$se["Robust", ]
data.frame(jump_hat = jump_hat, true_jump = true_jump,
           bias = jump_hat - true_jump,
           robust_ci_covers = abs(jump_hat - true_jump) <= 1.96 * se_hat)
```

Manipulation check (McCrary): the latent multiplier's density should be smooth
through the cutpoint — the algorithm doesn't sort sessions around it.

```{r density}
summary(rddensity(surge$running, c = 0))
```

# 5. Local elasticity, and where it lives on the demand curve

Convert the jump to an arc elasticity at the threshold — then be honest about
what it is: a **LATE**, valid at 1.25× base price, not the whole curve.

```{r curve}
p_lo <- base_p * (cut - step / 2); p_hi <- base_p * (cut + step / 2)
conv_lo <- lev$purchased[abs(lev$price - base_p * 1.2) < 1e-6]
conv_hi <- lev$purchased[abs(lev$price - base_p * 1.3) < 1e-6]
eps_local <- (jump_hat / mean(c(conv_lo, conv_hi))) /
             ((p_hi - p_lo) / mean(c(p_lo, p_hi)))
eps_local

truth_curve <- read.csv("../data/ground_truth_curves.csv")   # ORACLE
ggplot() +
  geom_line(data = truth_curve, aes(price, P_aggregate, color = "true curve (oracle)"),
            linewidth = 1) +
  geom_point(data = lev, aes(price, purchased, color = "conversion by surge level"),
             size = 2.5) +
  geom_vline(xintercept = base_p * cut, linetype = "dotted") +
  annotate("text", x = base_p * cut + 6, y = 0.45,
           label = "RD identifies THIS jump only") +
  labs(y = "P(buy)", color = NULL,
       title = "Step-priced conversion traces the curve; RD certifies one step of it")
```

# 6. Forecast (Phase 5): the decision at the margin

RD's forecast is a counterfactual, not a time series: expected demand per 1,000
sessions if the algorithm rounds *down* vs *up* at the cutpoint. This is
exactly the pricing decision the discontinuity licenses.

```{r forecast}
sessions_per_day <- 1000
data.frame(
  scenario = c("round down (price 30.0)", "round up (price 32.5)"),
  expected_units = sessions_per_day * c(conv_lo, conv_lo + jump_hat),
  truth          = sessions_per_day * c(conv_lo, conv_lo + true_jump)
)
```

Extrapolating this elasticity to prices far from 1.25× base is unlicensed — for
the whole curve you need the experiment (02) or a structural model. That
boundary between what the design identifies and what it doesn't is the point of
the exercise.
