---
title: "06 — Tobit / censored demand"
subtitle: "demand-sim · docs/03 §7, docs/02 · sales are not demand once inventory exists"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
```

`units_sold = min(units_demanded, on_hand)`. On sold-out days the observation
is right-censored: we know demand was *at least* sales. The observable
`sold_out` flag tells us which rows are censored — per-row censoring, so we use
`survival::survreg` directly (docs/03: `AER::tobit` wraps it for a fixed
threshold). Censoring concentrates on promo days here, so ignoring it biases
exactly the promo/price coefficients a pricing team cares about.

# 1. Data

```{r data}
library(survival)
library(ggplot2)

panel <- read.csv("../data/fact_sales_daily.csv")
panel$date <- as.Date(panel$date)
panel$series <- paste(panel$product_id, panel$store_id, sep = "_")
panel$log_price <- log(panel$avg_price)
panel$dow <- factor(format(panel$date, "%u"))
# pandas writes booleans as "True"/"False" — convert to logical explicitly
panel$promo_flag <- panel$promo_flag == "True"
panel$sold_out <- panel$sold_out == "True"

c(share_soldout_days = mean(panel$sold_out),
  share_soldout_promo = mean(panel$sold_out[panel$promo_flag]),
  share_soldout_nonpromo = mean(panel$sold_out[!panel$promo_flag]))
```

# 2. Train: naive GLM vs per-row-censored Tobit

Log-normal working model for daily demand. The Tobit likelihood treats
sold-out rows as `demand >= units_sold`; uncensored rows as observed.

```{r train}
panel$y <- log1p(panel$units_sold)
panel$uncensored <- as.numeric(!panel$sold_out)   # 1 = event observed

fit_naive <- lm(y ~ log_price + promo_flag + dow + series, data = panel)

fit_tobit <- survreg(Surv(y, uncensored, type = "right") ~
                       log_price + promo_flag + dow + series,
                     data = panel, dist = "gaussian")

data.frame(term = c("log_price", "promo_flagTRUE"),
           naive = coef(fit_naive)[c("log_price", "promo_flagTRUE")],
           tobit = coef(fit_tobit)[c("log_price", "promo_flagTRUE")])
```

The Tobit price and promo coefficients are larger in magnitude: the naive fit
learned from spikes that were clipped at the shelf.

# 3. Validate against the oracle

Predicted latent demand vs true `units_demanded`, split by censoring status —
the naive model should look fine on clean days and fail exactly on sold-out
days.

```{r validate}
mu_naive <- predict(fit_naive)
mu_tobit <- predict(fit_tobit)          # latent mean on log1p scale
panel$pred_naive <- expm1(mu_naive)
panel$pred_tobit <- expm1(mu_tobit + 0.5 * fit_tobit$scale^2)  # lognormal mean

by_status <- function(pred) {
  err <- pred - panel$units_demanded    # ORACLE target
  c(bias_all = mean(err),
    bias_soldout_days = mean(err[panel$sold_out]),
    bias_clean_days = mean(err[!panel$sold_out]))
}
rbind(naive = by_status(panel$pred_naive),
      tobit = by_status(panel$pred_tobit))
```

# 4. The demand curve it implies

For the highest-censoring series: fitted demand vs price under both models,
against true expected demand (`lambda_true`, oracle).

```{r curve}
cens <- aggregate(cbind(lost_sales, units_demanded) ~ series, panel, sum)
worst <- cens$series[which.max(cens$lost_sales / cens$units_demanded)]
s <- subset(panel, series == worst)

grid <- data.frame(log_price = log(seq(min(s$avg_price), max(s$avg_price), length.out = 40)),
                   promo_flag = FALSE, dow = factor(3, levels = levels(panel$dow)),
                   series = worst)
grid$price <- exp(grid$log_price)
grid$naive <- expm1(predict(fit_naive, grid))
grid$tobit <- expm1(predict(fit_tobit, grid) + 0.5 * fit_tobit$scale^2)

truth_pts <- aggregate(lambda_true ~ avg_price, s, mean)   # ORACLE

ggplot() +
  geom_line(data = grid, aes(price, naive, color = "naive (sales-trained)"),
            linetype = "dashed", linewidth = 1) +
  geom_line(data = grid, aes(price, tobit, color = "tobit (censoring-aware)"),
            linewidth = 1) +
  geom_point(data = truth_pts, aes(avg_price, lambda_true, color = "true mean demand (oracle)"),
             size = 3) +
  labs(x = "price", y = "units/day", color = NULL,
       title = paste("Series", worst, "— the naive curve is clipped at the shelf"))
```

The gap opens at the *low-price* end — promo prices are where stock ran out, so
the naive curve underestimates demand precisely where demand is largest.

# 5. Forecast (Phase 5): the R money table

Hold out the last 28 days; forecast with the naive fit, a drop-stockout-days
fit, and the Tobit fit; grade all three against oracle demand. This mirrors
docs/09's Python benchmark.

```{r forecast}
cutoff <- max(panel$date) - 27
tr <- subset(panel, date < cutoff); te <- subset(panel, date >= cutoff)

f_naive <- lm(y ~ log_price + promo_flag + dow + series, data = tr)
f_drop  <- lm(y ~ log_price + promo_flag + dow + series,
              data = subset(tr, !sold_out))
f_tobit <- survreg(Surv(y, uncensored, type = "right") ~
                     log_price + promo_flag + dow + series,
                   data = tr, dist = "gaussian")

pred <- data.frame(
  naive = expm1(predict(f_naive, te)),
  drop_stockouts = expm1(predict(f_drop, te)),
  tobit = expm1(predict(f_tobit, te) + 0.5 * f_tobit$scale^2)
)
money <- t(sapply(pred, function(p) {
  ed <- p - te$units_demanded; es <- p - te$units_sold   # oracle vs observable
  c(mae_vs_demand = mean(abs(ed)), rmse_vs_demand = sqrt(mean(ed^2)),
    bias_vs_demand = mean(ed), bias_vs_sales = mean(es))
}))
round(money, 3)
```

Read the last two columns together: the naive model's bias vs *sales* is small
— it looks calibrated on every metric a real forecaster can compute — while its
bias vs *demand* is systematically negative. That invisible gap is the thesis
of this project; the Tobit row closes most of it using only observable
information (`sold_out`, `units_sold`).
