---
title: "07 — IV / 2SLS and the control function"
subtitle: "demand-sim · docs/03 (parity table), docs/05 §3-§4 · rescuing the elasticity from endogenous prices"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
```

Same endogenous market as 04 (θ = −1.5, γ = 0.5): the pricer prices into
demand shocks, so OLS is attenuated. The classical fixes, in their reference
implementations: 2SLS via `AER::ivreg` (with the full diagnostic suite) and the
two-step control function. The identifying assumption for both: the cost
shifter `z` moves price and enters demand only through price (exclusion) —
true here by construction, an argument to be won anywhere else.

# 1. Data

```{r data}
library(AER)
library(ggplot2)

market <- read.csv("../data/endogenous_market.csv")
truth  <- read.csv("../data/truth_params.csv")
theta_true <- truth$value[truth$param == "theta_true"]
head(market)
```

# 2. The problem: naive OLS

```{r ols}
ols <- lm(log_q ~ log_p + x, data = market)
summary(ols)$coefficients["log_p", ]
```

~−1.1 against a truth of −1.5: every unit of pricing-into-shocks pushes the
estimate toward zero (`plim = theta + Cov(P,u)/Var(P)`, docs/05 §2).

# 3. Fix 1 — 2SLS

```{r iv}
iv <- ivreg(log_q ~ log_p + x | z + x, data = market)
summary(iv, diagnostics = TRUE)
```

Read the diagnostics, always: weak-instruments F (must clear ~10 by a lot),
Wu–Hausman (formally detects the endogeneity OLS ignored). Sargan is not
identified here — one instrument, one endogenous regressor, zero
overidentifying restrictions.

# 4. Fix 2 — Control function

First stage residual carries the endogenous part of price; including it in the
demand equation absorbs the confound. The t-test on `v_hat` IS the endogeneity
test. (This is the version that generalizes to the purchase logit — Petrin &
Train 2010.)

```{r cf}
first <- lm(log_p ~ z + x, data = market)
market$v_hat <- resid(first)
cf <- lm(log_q ~ log_p + x + v_hat, data = market)
summary(cf)$coefficients[c("log_p", "v_hat"), ]
```

# 5. Validate against the oracle

```{r validate}
res <- data.frame(
  estimator = c("OLS", "2SLS", "control function"),
  theta_hat = c(coef(ols)["log_p"], coef(iv)["log_p"], coef(cf)["log_p"]),
  se = c(summary(ols)$coefficients["log_p", 2],
         summary(iv)$coefficients["log_p", 2],
         summary(cf)$coefficients["log_p", 2])
)
res$bias <- res$theta_hat - theta_true
res$ci_covers_truth <- abs(res$bias) <= 1.96 * res$se
res
```

(The CF standard error is slightly optimistic — it ignores first-stage
estimation noise; bootstrap both steps for exact inference. 2SLS and CF point
estimates are identical in linear models, as they should be.)

# 6. The demand curve it implies

```{r curve}
p_bar <- mean(exp(market$log_p)); q_bar <- mean(exp(market$log_q))
grid <- seq(0.6 * p_bar, 1.6 * p_bar, length.out = 80)
curves <- rbind(
  data.frame(price = grid, q = q_bar * (grid / p_bar)^theta_true,       which = "truth (-1.5)"),
  data.frame(price = grid, q = q_bar * (grid / p_bar)^coef(ols)["log_p"], which = "OLS (attenuated)"),
  data.frame(price = grid, q = q_bar * (grid / p_bar)^coef(iv)["log_p"],  which = "2SLS")
)
ggplot(curves, aes(price, q, color = which)) + geom_line(linewidth = 1) +
  labs(y = "quantity", color = NULL,
       title = "The OLS curve says price cuts don't pay; the IV curve says they do")
```

# 7. Forecast (Phase 5): forecasting an intervention

The forecast that matters to a pricing team is counterfactual: what happens to
quantity if we *choose* to move price 10%? Under intervention the demand shock
does not come along for the ride — so the causal θ is the right multiplier,
and the OLS θ understates the response.

```{r forecast}
price_change <- 0.10
data.frame(
  theta_used = c("OLS", "2SLS", "truth"),
  pct_quantity_change = 100 * ((1 + price_change)^c(
    coef(ols)["log_p"], coef(iv)["log_p"], theta_true) - 1)
)

# and a plain holdout forecast, where the difference nearly vanishes:
n <- nrow(market); tr <- 1:floor(0.8 * n); te <- (floor(0.8 * n) + 1):n
gx <- lm(log_q ~ x, data = market[tr, ])
for_theta <- function(theta) {
  pred <- predict(gx, market[te, ]) + theta * (market$log_p[te] - mean(market$log_p[tr]))
  c(rmse = sqrt(mean((pred - market$log_q[te])^2)))
}
rbind(OLS = for_theta(coef(ols)["log_p"]),
      `2SLS` = for_theta(coef(iv)["log_p"]),
      truth = for_theta(theta_true))
```

The holdout table is the trap in miniature: under the *observational* price
process, the wrong θ predicts almost as well — because prices keep co-moving
with shocks. Only under intervention (the first table) does the bias surface,
as a ~4-point understatement of the quantity response. Prediction accuracy is
not a test of causal validity; that is why this project scores against an
oracle instead.
